Pythran: Enabling Static Optimization of Scientific Python Programs

### 2014-01-30 doc/papers/iop2014/paper.tex
update paper

    % FIXME  (explain that annotations can be added to the code?)
    Unlike the Cython approach , Pythran input code remains compatible with the
    Python interpreter. Output code is generally as efficient as the annotated
    Cython equivalent, if not more, but without the backward compatibility
    loss.

@@

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high. % FIXME cut sentence It is partly due to the interpration cost,
but also to Python's dynamic binding. Additionally, the Python compiler does
not perform any optimization on the byte code, while scientific applications
are first-class candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The goal of an optimizing compiler % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

@@

Except for PyPy, none of these compilers apply any of the static optimization
techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled language such as C or C++.  Translators to
statically compiled languages do take advantage of them indirectly, but the
quality of the generated code may prevent advanced optimizations. For instance
GCC hardly vectorizes the output of Cython, while the array semantic is
avaialable at the Python level. Taking into account the specificities of the
Python language can unlock many new transformations.  For instance, PyPy
automatically converts the \texttt{range} builtin into \texttt{xrange} by using
a dedicated structure called \texttt{range-list}.

@@

Pythran IR is similar to the Python AST, as defined in the \texttt{ast} module, except
that several nodes are forbidden (for instance because Pythran does not support
user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
to others to form a simpler and easier to deal with AST for further analyses and
optimizations. The transformations applied by Pythran on Python AST are the
following:

@@

    \item all imports are fully expanded to make functions and global variables access paths explicit,

@@

    \item compiling the generated C++ code into native code.

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its life
time, but it cannot be assigned to a string if it has already been assigned to
an integer.  As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected. This technique is
very similar to Python's duck typing, except that it is checked at compile
time, as illustrated by the following implementation of a generic dot product
in Python:

@@

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions called in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverage the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

@@

Without this kind of information, even a simple instruction like
\texttt{sum(a)} would yield very few information as there is no guarantee that
the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
previously stated that \texttt{sum = len}.

When turning Python AST into Pythran IR, nested functions are turned into
global functions taking their expanded closure as parameter. This closure is
computed using the information provided by the \texttt{Globals} and
\texttt{ImportedIds} analyse. The former statically computes the state of the
dictionary of globals, and the latter computes the set of identifiers used by
an instruction but not declared in this instruction. For instance in the
following snippet:

@@

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because \texttt{print b} involves an \textbf{I/O}.
% Which is not really usefull as there is no way to say pythran we want to execut it in parallel
% (but logging information is still displayed)

@@

However, there are some information available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functor with an
internal state and a goto dispatch table to represent generators. This approach
is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]i + i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i] + i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
% PB : Not sure it is needed as audiance don't really care about it.

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developers to
be able to take advantage of them.
%Does python programmer really use it? Or people using it may write their native code?
% They may write it using python only to prototyping?

@@

Table~\ref{tbl:nqueen} illustrates the benchmark results for CPython, Cython,
PyPy and Pythran.  It showcases the fact that Pythran benefits a lot from
high-level transformations, for instance \emph{lazy evaluation}, which prevents
the many allocation / deallocation triggered by Cython and PyPy runs.  A deeper
look at the Pythran profiling trace shows that more than half of the execution
time is spent allocating and deallocating a \texttt{set} used in the internal
loop. There is a memory allocation invariant that could be taken advantage of
there, but none of the compiler does so.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expression is still note mature for numpy expression, but there is a
strong on-going effort of the community to fix this.

@@

The ``growcut'' application is an image processing code where all array indexing
are explicit. In that sense, this benchmark is complementary of the previous
one, as it evaluates how well explicit array access are compiled.
Table~\ref{tbl:growcut} summarizes the results. All compiler yield impressive
speedups on that benchmark, but this is is merely due to the fact that
accessing individual elements is extremely slow in numpy, and the direct C
translation of the Python code is very close to what a C developer would write.
Pythran appears to outperforms Cython once all access are explicit, most
certainly due to cleaner code generation.

@@

The ``blacksholes'' application is a benchmarks that showcases the
translation cost from the Python space to native space. Indeed we forced the
use of a \texttt{list} as input of the kernel instead of a numpy array. As a
consequence there is a costly translation step from a structure that contains
an array of pointers to generic Python objects to a dense structure involved.
Results are shown in Table~\ref{tbl:black}. They show that conversion code is
less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
it does not suffer from a similar conversion penalty.

@@

To conclude, limiting Python to a statically typed subset does not hinders the
expressiveness when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performances of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to
write for the average programmer.

@@

This project has been partially funded by the CARP Project and the SILKAN
Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
their contributions to development of the Pythran compiler.

>>>>

    % FIXME  (explain that annotations can be added to the code?)
    Unlike the Cython approach, code fed to Pythran remains compatible with the
    Python interpreter. The result is generally as efficient as Cython with
    the equivalent annotated Cython code, if not more, 
    but without the backward compatibility loss.

@@

However, the overhead of running a scientific application written in Python
compared to the same algorithm written in a statically compiled language such
as C is high. % FIXME cut sentence 
It is partly due to the interpration cost,
but also because of Python's dynamic binding. Additionally, the Python compiler does
not perform any optimization on the byte code, while scientific applications
are first-class candidates for many of them.

Since scientific applications are often told to spend 90\% of their time in
10\% of the code, it is natural to focus on those computation-intensive pieces 
of code. The goal of a specialized compiler % FIXME (of what) 
may not be to optimize the full Python application, but rather a subset of the application.

@@

Except for PyPy, none of these compilers apply any of the static optimization
techniques that have been known and successfully applied 
% FIXME And Numba? It generates LLVM, which uses static optimization techniques
for decades to statically compiled languages such as C or C++.  Translators to
statically compiled languages do take advantage of them indirectly, but the
quality of the generated code may prevent advanced optimizations. For instance
GCC hardly vectorizes the output of Cython, while the array semantic is
avaialable at the Python level. Taking into account the specificities of the
Python language can unlock many new transformations.  For instance, PyPy
automatically converts the \texttt{range} builtin into \texttt{xrange} by using
a dedicated structure called \texttt{range-list}.

@@

Pythran IR is similar to the Python AST, as defined in the \texttt{ast} module, except
that several nodes are forbidden (for instance because Pythran does not support
user-defined classes, or the \texttt{exec} instruction), and some nodes are converted
to others to form a simpler and easier to deal with AST for further analyses and
optimizations. The transformations applied by Pythran on the Python AST are the
following:

@@

    \item all imports are fully expanded to make the access paths of functions and global variables explicit,

@@

    \item compiling the generated C++ code into native code.%FIXME maybe native module?

@@

The first step requires to map polymorphic variables and polymorphic functions
from the Python world to C++. Pythran only supports polymorphic variables for
functions, i.e. a variable can hold several function objects during its lifetime, 
but it cannot be assigned to a string if it has already been assigned to
an integer.  As shown later, it is possible to detect several false variable
polymorphism cases using use-def chains. Function polymorphism is achieved
through template parameters: a template function can be applied to several
types as long as an implicit structural typing is respected. This technique is
very similar to Python's duck typing, except that it is checked at compile
time, as illustrated by the following implementation of a generic dot product
in Python:

@@

The second step only consists in the instantiation of the top-level functions of
the module, using user-provided signatures. Template instantiation then triggers
the different correctly typed instantiations for all functions called in the
module. Note that the user only needs to provide the type of the functions
exported outside the module. The possible types of all internal functions are
then inferred from the call sites. Note that this complex step leverages the 
C++11 compiler capabilities: Pythran generates only one line using user-provided
signature to trigger the template instanciation by the C++ compiler.

@@

Without this kind of information, even a simple instruction like
\texttt{sum(a)} would yield very little information as there is no guarantee that
the \texttt{sum} identifiers point to the \texttt{sum} built-in, as it could have been 
previously stated that \texttt{sum = len}.

When turning Python AST into Pythran IR, nested functions are turned into
global functions taking their expanded closure as parameter. This closure is
computed using the information provided by the \texttt{Globals} and
\texttt{ImportedIds} analyses. The former statically computes the state of the
dictionary of globals, and the latter computes the set of identifiers used by
an instruction but not declared in this instruction. For instance in the
following snippet:

@@

Several analyses depend on the \texttt{PureFunctions} analysis.
\texttt{ParallelMaps} uses aliasing information to check if an identifier
points to the \texttt{map} intrinsic, and checks if the first argument is a
pure function using \texttt{PureFunctions}. In that case the \texttt{map} is
added to the set of parallel maps, because it can be executed in any order.
This is the case for the first \texttt{map} in the following snippet, but not
for the second because \texttt{print b} involves an \textbf{I/O}.
% Which is not really usefull as there is no way to say pythran we want to execute it in parallel
% (but logging information is still displayed)

@@

However, there is some information available at the Python level that cannot
be recovered at the C++ level. For instance, Pythran uses functors with an
internal state and a goto dispatch table to represent generators. This approach
is not very efficient, especially for trivial cases.
Such trivial cases appear when a generator expression is converted, in the
front-end, to a looping generator. To avoid this extra cost, Pythran turns
generator expressions into call to \texttt{imap} and \texttt{ifilter} from the
\texttt{itertools} module whenever possible, removing the unnecessary goto
dispatching table. This kind of transformation cannot be made by the C++
compiler. For instance, the one-liner \texttt{len(set(vec[i]i + i for i in cols))}
extracted from the \texttt{nqueens} benchmarks from the Unladen Swallow project
is rewritten as \texttt{len(set(itertools.imap(lambda i: vec[i] + i,cols)))}.
This new form is less efficient in pure Python (it implies one extra function
call per iteration), but can be compiled into C++ more efficiently than a
generic generator.

@@

When benchmarked with the \texttt{timeit} module on an hyper-threaded quad-core 
i7, the pure Python execution yields: % FIXME sandy-brige? nehalem?...
% PB : Not sure it is needed as audience doesn't really care about it.

@@

Many scientific applications can benefit from the parallel execution of their
kernels. As modern computers generally feature several processors and several
cores per processor, it is critical for the scientific application developers to
be able to take advantage of them.
%Do python programmers really use it? Or people using it may write their native code?
% They may write it using python only to prototyping?

@@

Table~\ref{tbl:nqueen} illustrates the benchmark results for CPython, Cython,
PyPy and Pythran.  It showcases the fact that Pythran benefits a lot from
high-level transformations, for instance \emph{lazy evaluation}, which prevents
the many allocation / deallocation triggered by Cython and PyPy runs.  A deeper
look at the Pythran profiling trace shows that more than half of the execution
time is spent allocating and deallocating a \texttt{set} used in the internal
loop. There is a memory allocation invariant that could be taken advantage of
there, but none of the compilers do so.

@@

The ``rosen der'' kernel computes a complex numpy expression that accesses
data through a one dimensional slice. It is interesting to evaluate how much of
the expression can be compiled to regular, efficient loop while avoiding
temporaries.  Table~\ref{tbl:rosen} illustrate the benchmark results for
CPython, Cython, PyPy and Pythran. Cython slightly outperforms Pythran on that
test case, but Cython code had to be manually transformed into an explicit
loop. Moreover, activating automatic parallelization for Pythran yields an
additional $\times4$ speedup without a change to the input, while Cython needs
manual code annotation to achieve a similar result. It seems that PyPy support
for numpy expressions is still not mature, but there is a
strong ongoing effort of the community to fix this.

@@

The ``growcut'' application is an image processing code where all array indexing
is explicit. In that sense, this benchmark complements the previous
one, as it evaluates how well explicit array access are compiled.
Table~\ref{tbl:growcut} summarizes the results. Every compiler yield impressive
speedups on that benchmark, but this is is merely due to the fact that
accessing individual elements is extremely slow in numpy, and the direct C
translation of the Python code is very close to what a C developer would write.
Pythran appears to outperforms Cython once all access are explicit, most
certainly due to cleaner code generation.

@@

The ``blacksholes'' application is a benchmark that showcases the
translation cost from the Python space to native space. Indeed we forced the
use of a \texttt{list} as input of the kernel instead of a numpy array. As a
consequence there is a costly translation step from a structure that contains
an array of generic Python objects pointers to a dense structure.
Results are shown in Table~\ref{tbl:black}. They show that that conversion is
less efficient in Cython than in Pythran. PyPy also outperforms Cython, because
it does not suffer from that problem.

@@

To conclude, limiting Python to a statically typed subset does not hinder the
expressiveness when it comes to scientific or mathematic computations, but makes
it possible to use a wide variety of classical optimizations to help Python
match the performances of statically compiled language. Moreover, one can use
high level information to generate efficient code that would be difficult to
write for the average programmer.

@@

This project has been partially funded by the CARP Project and the SILKAN
Company.  The authors would like to thank Yuancheng Peng and Eliott Coyac for
their contributions to the development of the Pythran compiler.
